Title Goes Here

### 2021-04-09 analysis/paper/shceof.Rmd
Promedia ponderando cantidad de días
Closes #23
Un montón de laburo y bastante al pedo porque (naturalmente) no cambió casi nada.

```{r read-indices}
enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
  as.data.table() %>%
  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
  na.omit() %>%
  .[season(time) == "SON"] %>%
  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]

sams <- readRDS(data_path("derived", "sam.Rds")) %>%
  .[season(time) == "SON"] %>%
  .[, dim :=  lubridate::days_in_month(time)] %>%
  .[, .(estimate = weighted.mean(estimate, dim)),
    by = .(lev, term, time = seasonally(time))]

@@

```{r read-data}
era5 <- ReadNetCDF(ERA5(), vars = c(hgt = "z",
                                    t = "t"),
                   subset = list(level = list(850, 700, 200, 50),
                                 latitude = -90:10)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt), 
        t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
```

```{r compute-psi}
psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(vo = mean(vo)), by = .(lon, lat, lev, time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  setnames("value", "psi") %>% 
  .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
```

```{r o3}
o3 <- O3() %>% 
  ReadNetCDF(vars = c(toc = "tco3")) %>% 
  normalise_coords() %>% 
  .[, lon := ConvertLongitude(lon)] %>% 
  .[lat <= -20] %>% 
  .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]

@@

```{r psa}
psa <- ReadNetCDF(ERA5(), vars = c(hgt = "z"),
                   subset = list(level = list(500),
                                 latitude = -90:0)) %>% 
  normalise_coords() %>% 
  .[, .(hgt = mean(hgt)), by = .(lon, lat, time = seasonally(time))] %>% 
  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>%
  EOF(value ~ time | lon + lat, n = 1:3, data = .)

@@

```{r compute-ceof}
invisible(
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = -90:10,
                           level = list(200, 50))) %>%
    normalise_coords() %>%
    .[season(time) == "SON"] %>%
    .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
    .[, season := season(time)] %>%
    {
      ceof <<- .[, .(eof = list(compute_ceof(.SD))), by = .(season)]
      ceof_splitted <<-  .[, levs := lev] %>%
        .[, .(eof = list(compute_ceof(.SD, n = 1:3))), by = .(season, levs)]

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]

@@

```{r t}
t <- ERA5() %>% 
  ReadNetCDF(vars = c("t", hgt = "z", "o3"), 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, .(t = mean(t),
        o3 = mean(o3),
        hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))]
```

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value. To this synthetic dataset we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a). The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

@@

```{r sst}
sst <- ERA5_SST() %>%
  ReadNetCDF(vars = c("t" = "sst")) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(t = mean(t)), by = .(lon, lat, time = seasonally(time))]
```

@@

gdata %>% 
  ggplot(aes(arg, oni)) +
  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
   annotate(shadowtext:::GeomShadowText,
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
           color = "gray60", bg.colour = "white", bg.r = 0.2,
           size = 3,
           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +

@@

```{r cmap}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, .(pp = mean(pp, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
```

@@

gdata %>% 
  ggplot(aes(Arg(hgt), pp)) +
  geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
  annotate(shadowtext:::GeomShadowText,
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
           color = "gray60", bg.colour = "white", bg.r = 0.2,
           size = 3,
           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +

>>>>

```{r read-indices}
enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
  as.data.table() %>%
  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
  na.omit() %>%
  .[season(time) == "SON"] %>%
  .[, oni := oni*season_weights(time)] %>% 
  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]

sams <- readRDS(data_path("derived", "sam.Rds")) %>%
  .[season(time) == "SON"] %>%
  .[, estimate := estimate*season_weights(time)] %>% 
  .[, .(estimate = mean(estimate)),
    by = .(lev, term, time = seasonally(time))]

@@

```{r read-data}
era5 <- ReadNetCDF(ERA5(), vars = c(hgt = "z",
                                    t = "t"),
                   subset = list(level = list(850, 700, 200, 50),
                                 latitude = -90:10)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w), 
        t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))]
```

```{r compute-psi}
psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(vo = mean(vo*w)), by = .(lon, lat, lev, time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  setnames("value", "psi") %>% 
  .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
```

```{r o3}
o3 <- O3() %>% 
  ReadNetCDF(vars = c(toc = "tco3")) %>% 
  normalise_coords() %>% 
  .[, lon := ConvertLongitude(lon)] %>% 
  .[lat <= -20] %>% 
  .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
  .[, w := season_weights(time)] %>% 
  .[, .(toc = mean(toc*w)), by = .(lon, lat, time = seasonally(time))]

@@

```{r psa}
psa <- ReadNetCDF(ERA5(), vars = c(hgt = "z"),
                   subset = list(level = list(500),
                                 latitude = -90:0)) %>% 
  normalise_coords() %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))] %>% 
  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>%
  EOF(value ~ time | lon + lat, n = 1:3, data = .)

@@

```{r compute-ceof}
invisible(
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = -90:10,
                           level = list(200, 50))) %>%
    normalise_coords() %>%
    .[season(time) == "SON"] %>%
    .[, w := season_weights(time)] %>% 
    .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
    .[, season := season(time)] %>%
    {
      ceof <<- .[, .(eof = list(compute_ceof(.SD))), by = .(season)]
      ceof_splitted <<-  .[, levs := lev] %>%
        .[, .(eof = list(compute_ceof(.SD, n = 1:3))), by = .(season, levs)]

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]

@@

```{r t}
t <- ERA5() %>% 
  ReadNetCDF(vars = c("t", hgt = "z", "o3"), 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(t   = mean(t*w),
        o3  = mean(o3*w),
        hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
```

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value. To this synthetic dataset we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a). The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

@@

```{r sst}
sst <- ERA5_SST() %>%
  ReadNetCDF(vars = c("t" = "sst")) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, w := season_weights(time)] %>% 
  .[, .(t = mean(t*w)), by = .(lon, lat, time = seasonally(time))]
```

@@

gdata %>% 
  ggplot(aes(arg, oni)) +
  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
   annotate(shadowtext:::GeomShadowText,
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
           color = "gray60", bg.colour = "white", bg.r = 0.2,
           size = 3,
           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +

@@

```{r cmap}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, w := season_weights(time)] %>% 
  .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
```

@@

gdata %>% 
  ggplot(aes(Arg(hgt), pp)) +
  geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
  annotate(shadowtext:::GeomShadowText,
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
           color = "gray60", bg.colour = "white", bg.r = 0.2,
           size = 3,
           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
